---
title: "lab7"
format: html
---

📌 Project Title

Evaluating Which Human In-Vitro Cell Model Best Reflects COVID-19 Lung Biology Using RNA-seq Data

⸻

📌 Research Question

Among the human cell models infected with SARS-CoV-2 in GSE147507 (NHBE, A549, A549-ACE2, and Calu-3), which one shows gene expression changes most similar to real COVID-19 human lung tissue?

Put even more simply:

Which cell line behaves most like actual COVID-19 lungs?

⸻

📌 Rationale (Why this project matters)

1. Why compare cell lines to real lungs?

Scientists use cell lines to model COVID-19 in the lab. However:
	•	Not all cell lines respond to infection the same way.
	•	Some cell lines barely get infected.
	•	Others have immune responses drastically different from real human lungs.

To know which model is “good,” we can compare:

cell line response vs real COVID-19 lung tissue response

using gene expression changes after infection.

This is called model fidelity, meaning “how well an experimental model represents real human disease.”

⸻

2. What is RNA-seq differential expression (DE)?

RNA-seq measures how much each gene is expressed in a sample.

Differential expression means:
	•	Compare two conditions
(e.g., infected vs mock)
	•	Identify genes that go up or down during infection
	•	This tells us what biological pathways are activated

In this project you will:
	•	Compute DEGs for each cell line (infected vs mock)
	•	Compute DEGs for lung tissue (COVID vs healthy)
	•	Compare the results

⸻

3. Why PCA? (Simple definition)

PCA = Principal Component Analysis.
	•	It reduces thousands of genes into 2 dimensions.
	•	Samples that are more similar cluster closer together.
	•	Samples that are different appear farther apart.

In this project, PCA will show:

which cell line’s infected samples cluster closest to real COVID-19 lung samples.

This is visual evidence of similarity.

⸻

4. Why this project fits biomedical informatics

This project uses classic BI skills:
	•	Data exploration
	•	PCA visualization
	•	Multi-condition comparison
	•	Quantitative similarity metrics
	•	Translational modeling (in vitro → in vivo)

It’s meaningful, feasible in one week, and easy to interpret.

⸻

📌 Dataset Summary (for memory)

You will use ONLY human samples:

Human Cell Lines (in vitro)
	•	NHBE (primary airway cells)
	•	A549
	•	A549-ACE2 (modified with ACE2)
	•	Calu-3

→ each has mock and SARS-CoV-2 infected samples.

Human Lung Tissue (real patients)
	•	Healthy lung biopsies
	•	COVID-19 lung biopsies

→ comparison = real disease signature.

⸻

📌 Project Hypothesis (simple)

Calu-3 and NHBE cells will show gene expression changes more similar to COVID-19 patient lungs than A549 or A549-ACE2 cells.

Because:
	•	Calu-3 and NHBE are naturally permissive (ACE2-expressing)
	•	A549 is minimally permissive unless modified

⸻

🧭 📌 Basic Step-by-Step Outline (Beginner Friendly)

(This is the part you’ll follow while doing the project.)

⸻

STEP 0 — Load data & metadata
	•	Import counts matrix
	•	Import sample info (cell line, mock/infected, etc.)
	•	Convert into DESeq2 dataset
	
```{r}
# loading in packages
library(tidyverse) # already know about this from Step 1 script
library(edgeR) # well known package for differential expression analysis, but we only use for the DGEList object and for normalization methods
library(matrixStats) # let's us easily calculate stats on rows or columns of a data matrix
library(cowplot)
library(DescTools)
library(tximport)
library(reshape2)
library(edgeR)
library(matrixStats)
library(cowplot)
library(plotly)
```

	
```{r}
# loading the study data
studyDesign <- read.table("~/MBMI/CAMB7140/labs/lab7/studyDesign.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>% 
  as.tibble()

# getting sample labels 
sampleLabels <- studyDesign$sample

# importing human raw reads counts data
human_raw_reads <- read_tsv('~/MBMI/CAMB7140/labs/lab7/GSE147507_RawReadCounts_Human.tsv')

# renaming the first column 
names(human_raw_reads)[1] <- "geneID"

# converting raw reads to matrix
human_raw_reads <- human_raw_reads %>% 
  column_to_rownames("geneID") %>% 
  as.matrix()

# creating a DGEList object from the raw reads counts data for humans 
human_DGEList <- DGEList(human_raw_reads)
```

```{r}
# use the 'cpm' function from EdgeR to get counts per million
cpm <- edgeR::cpm(human_DGEList) 
colSums(cpm)
log2.cpm <- edgeR::cpm(human_DGEList, log=TRUE)
```

```{r}
# 'coerce' your data matrix to a dataframe so that you can use tidyverse tools on it
log2.cpm.df <- as_tibble(log2.cpm, rownames = "geneID")
log2.cpm.df
# use the tidy package to 'pivot' your dataframe (from wide to long)
log2.cpm.df.pivot <- pivot_longer(log2.cpm.df, # dataframe to be pivoted
                                  cols = Series1_NHBE_Mock_1:`Series16_A549-ACE2_SARS-CoV-2_Rux_3`, # column names to be stored as a SINGLE variable
                                  names_to = "samples", # name of that new variable (column)
                                  values_to = "expression") # name of new variable (column) storing all the values (data)

# let's look at the impact of pivoting the data
log2.cpm.df.pivot
```

```{r}
# filtering 
cpm <- cpm(human_DGEList)
keepers <- rowSums(cpm>1)>=5 #user defined
human_DGEList.filtered <- human_DGEList[keepers,]

log2.cpm.filtered <- cpm(human_DGEList.filtered, log=TRUE)
log2.cpm.filtered.df <- as_tibble(log2.cpm.filtered, rownames = "geneID")
colnames(log2.cpm.filtered.df) <- c("geneID", sampleLabels)

# normalizing
human_DGEList.filtered.norm <- calcNormFactors(human_DGEList.filtered, method = "TMM")
log2.cpm.filtered.norm <- cpm(human_DGEList.filtered.norm, log=TRUE)
log2.cpm.filtered.norm.df <- as_tibble(log2.cpm.filtered.norm, rownames = "geneID")
colnames(log2.cpm.filtered.norm.df) <- c("geneID", sampleLabels)
log2.cpm.filtered.norm.df.pivot <- pivot_longer(log2.cpm.filtered.norm.df, # dataframe to be pivoted
                                                cols = -1, # column names to be stored as a SINGLE variable
                                                names_to = "samples", # name of that new variable (column)
                                                values_to = "expression") # name of new variable (column) storing all the values (data)
```

⸻

STEP 1 — Run DESeq2 for each cell line

For each of the 4 cell types:
	•	NHBE: infected vs mock
	•	A549: infected vs mock
	•	A549-ACE2: infected vs mock
	•	Calu-3: infected vs mock


```{r}
run_dge_analysis <- function(cell_line){
  
  if (cell_line %in% c("primary NHBE", "A549", "A549-ACE2", "Calu-3")) {
    studyDesign <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(treatment %in% c('Mock treatment', 'SARS-CoV-2 infected (MOI 2)')) %>% 
      mutate(group = case_when(treatment == 'Mock treatment' ~ "control",
                                 treatment == 'SARS-CoV-2 infected (MOI 2)' ~ "infected"))
  } else if (cell_line == "Lung Biopsy") {
     studyDesign <- studyDesign %>% 
       filter(host_common_name == "human") %>% 
       filter(tissue_cell_type == "Lung Biopsy") %>% 
       mutate(group = case_when(tolower(subject_status) %like% 'healthy%' ~ "control",
                           tolower(subject_status) %like% 'covid%' ~ "infected"))
  } else {
    stop("Invalid cell line. Please choose from: primary NHBE, A549, A549-ACE2, Calu-3, lung_biopsy.")
  }
  
  if (cell_line == "primary NHBE") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(cell_line == 'primary NHBE') %>% 
      filter(group %in% c('control', 'infected'))
  } else if (cell_line == "A549") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(cell_line == 'A549') %>% 
      filter(group %in% c('control', 'infected'))
  } else if (cell_line == "A549-ACE2") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(cell_line == 'A549-ACE2') %>% 
      filter(group %in% c('control', 'infected'))
  } else if (cell_line == "Calu-3") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(cell_line == 'Calu-3') %>% 
      filter(group %in% c('control', 'infected'))
  } else if (cell_line == "Calu-3") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(cell_line == 'Calu-3') %>% 
      filter(group %in% c('control', 'infected'))
  } else if (cell_line == "Lung Biopsy") {
    targets <- studyDesign %>% 
      filter(host_common_name == "human") %>% 
      filter(tissue_cell_type == "Lung Biopsy") %>% 
      filter(group %in% c('control', 'infected'))
  }
  
  samples <- targets$sample
  group <- factor(targets$group)
  design <- model.matrix(~0 + group)
  colnames(design) <- levels(group)
  
  v.DEGList.filtered.norm <- voom(human_DGEList.filtered.norm[ , samples ], design, plot = FALSE)
  fit <- lmFit(v.DEGList.filtered.norm, design)
  contrast.matrix <- makeContrasts(infection = infected - control,
                                   levels=design)
  
  fits <- contrasts.fit(fit, contrast.matrix)
  ebFit <- eBayes(fits)
  myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")
  myTopHits.df <- myTopHits %>%
    as_tibble(rownames = "geneID")
  
  vplot <- ggplot(myTopHits.df) +
    aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", geneID)) +
    geom_point(size=2) +
    geom_hline(yintercept = -log10(0.01), linetype="longdash", colour="grey", linewidth=1) +
    geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", linewidth=1) +
    geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", linewidth=1) +
    #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
    #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
    labs(title="Volcano plot",
         subtitle = "SARS-CoV-2 infected vs mock",
         caption=paste0("produced on ", Sys.time())) +
    theme_bw()
  
  vplot_ggplotly <- ggplotly(vplot)
  
  
  results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.01, lfc=2)
  colnames(v.DEGList.filtered.norm$E) <- samples
  diffGenes <- v.DEGList.filtered.norm$E[results[,1] !=0,]
  diffGenes.df <- as_tibble(diffGenes, rownames = "geneID")
  
  return(list("DESeq2_results" = myTopHits.df,
              "volcano_plot" = vplot_ggplotly,
              "DEGs" = diffGenes.df))
}


```

```{r}
# comparing upregulated counts in lung biopsy samples to upregulated counts in cell lines
# lung upregulated genes

run_upregulated_share <- function(cell_lines = c("primary NHBE", "A549", "A549-ACE2", "Calu-3")){
  
  lung_upregulated_genes <- run_dge_analysis("Lung Biopsy")$DESeq2_results %>% 
  filter(logFC >= 0) %>% 
  filter(adj.P.Val <= 0.05) %>% 
  arrange(desc(logFC)) %>% 
  arrange(adj.P.Val)
  
  cell_line_upregulated_list <- list()
  lung_cell_line_shared_upregulated_list <- list()
  
  for (cell_line in cell_lines) {
    cell_line_upregulated_genes <- run_dge_analysis(cell_line)$DESeq2_results %>% 
      filter(logFC >= 0) %>% 
      filter(adj.P.Val <= 0.05) %>% 
      arrange(desc(logFC)) %>% 
      arrange(adj.P.Val)
    
    cell_line_upregulated_list[[cell_line]] <- cell_line_upregulated_genes
    
    shared_upregulated_genes <- lung_upregulated_genes %>% select(geneID) %>%
      inner_join(cell_line_upregulated_list[[cell_line]] %>% select(geneID), by = "geneID") %>% 
      mutate(shared_upregulated = n()) %>% 
      distinct(shared_upregulated) %>% 
      mutate(cell_line = cell_line,
             pct_of_all_lung_upregulated = (shared_upregulated/lung_upregulated_genes %>% summarize(count = n()) %>% pull()) * 100) %>% 
      select(cell_line, shared_upregulated, pct_of_all_lung_upregulated)
    
    lung_cell_line_shared_upregulated_list[[cell_line]] <- shared_upregulated_genes
    
  }
  
  combined_results <- dplyr::bind_rows(lung_cell_line_shared_upregulated_list)
  
  return(list(
      cell_line_upregulated = cell_line_upregulated_list,
      shared_counts = lung_cell_line_shared_upregulated_list,
      combined_results = combined_results
    ))
}

# Load ggplot2
library(ggplot2)

# Create data
dge_results <- run_upregulated_share()$combined_results

# Barplot
dge_plot <- ggplot(
  dge_results %>% mutate(cell_line = factor(cell_line)),
  aes(y = cell_line, x = pct_of_all_lung_upregulated, fill = cell_line)
) +
  geom_col() +
  xlim(0, 100) +
  labs(
    title = "Percentage of Lung Upregulated Genes Shared with Cell Lines",
    x = "Percentage of Lung Upregulated Genes",
    y = "Cell Line"
  ) +
  theme_light() +
  theme(legend.position = "none")
```


Produce:
	•	DESeq2 results table
	•	Volcano plot
	•	List of significant DEGs
	•	(Optional) enriched pathways

You now have 4 cell-line infection signatures.

⸻

STEP 2 — Run DESeq2 for lung tissue

Compare:
	•	COVID lung vs healthy lung

Produce:
	•	DESeq2 results table
	•	Volcano plot
	•	List of DEGs

You now have the “real lung disease signature.”

⸻

STEP 3 — Compare cell lines to lung tissue

3A. Fold-change correlation

For each cell line:
	•	Compute correlation of |log2FC| with lung |log2FC|

Higher correlation = more similar to real lungs.

3B. DEG overlap

Count how many up-regulated and down-regulated genes overlap with lungs.

3C. Pathway overlap

Compare enriched pathways (e.g., interferon, cytokines, apoptosis).

⸻

STEP 4 — PCA

Run PCA on all human samples:
	•	Healthy lungs
	•	COVID lungs
	•	Mock cell lines
	•	Infected cell lines

Look for:

👉 The infected cell line samples that cluster closest to the COVID lung samples.

This is visual similarity evidence.


```{r}
library(dplyr)
library(stringr)
library(tibble)
library(ggplot2)

############################################################
# 1. Filter metadata to relevant samples
############################################################

filtered_studyDesign <- studyDesign %>% 
  filter(host_common_name == "human") %>%
  filter(cell_line %in% c("primary NHBE", "A549", "A549-ACE2", "Calu-3") |
         tissue_cell_type == "Lung Biopsy") %>%
  filter(
    treatment %in% c("Mock treatment", "SARS-CoV-2 infected (MOI 2)") |
    str_detect(tolower(subject_status), "healthy") |
    str_detect(tolower(subject_status), "covid")
  ) %>%
  mutate(
    group = case_when(
      treatment == "Mock treatment" ~ "control",
      treatment == "SARS-CoV-2 infected (MOI 2)" ~ "infected",
      str_detect(tolower(subject_status), "healthy") ~ "control",
      str_detect(tolower(subject_status), "covid") ~ "infected"
    )
  ) %>%
  mutate(
    cell_line = case_when(
      tissue_cell_type == "Lung Biopsy" ~ "Lung Biopsy",
      TRUE ~ cell_line
    )
  )

############################################################
# 2. Select samples used in PCA
############################################################

selected_samples <- filtered_studyDesign$sample

############################################################
# 3. Subset normalized expression matrix
############################################################

expr_matrix <- log2.cpm.filtered.norm[, selected_samples]

############################################################
# 4. Remove housekeeping genes using variance filtering
############################################################

gene_var <- apply(expr_matrix, 1, var)

top_genes <- names(sort(gene_var, decreasing = TRUE))[1:1000]

pca_matrix <- expr_matrix[top_genes, ]

############################################################
# 5. Run PCA
############################################################

pca.res <- prcomp(
  t(pca_matrix),
  scale. = TRUE,
  retx = TRUE
)

############################################################
# 6. Compute variance explained
############################################################

pc.var <- pca.res$sdev^2
pc.per <- round(pc.var / sum(pc.var) * 100, 1)

############################################################
# 7. Convert PCA output to dataframe and attach metadata
############################################################

pca.res.df <- as_tibble(pca.res$x, rownames = "sample") %>%
  left_join(
    filtered_studyDesign %>%
      mutate(model_condition = paste(cell_line, group, sep = "_")) %>%
      select(sample, cell_line, group, model_condition),
    by = "sample"
  )

############################################################
# 8. PCA plot
############################################################

pca.plot <- ggplot(pca.res.df) +
  aes(
    x = PC1,
    y = PC2,
    color = cell_line,
    shape = group
  ) +
  geom_point(size = 4) +
  stat_ellipse() +
  xlab(paste0("PC1 (", pc.per[1], "%)")) +
  ylab(paste0("PC2 (", pc.per[2], "%)")) +
  labs(
    title = "PCA plot (Top 1000 most variable genes)"
  ) +
  coord_fixed() +
  theme_bw() + 
  guides(
  color = guide_legend(nrow = 1),
  shape = guide_legend(nrow = 1)
  ) + 
  theme(
  legend.position = "bottom",
  legend.box = "vertical"
)

pca.plot
```

```{r}
dge_plot
ggplotly(pca.plot)
```
```{r}
library(patchwork)

pca_sub <- pca.plot +
  labs(title = NULL)

dge_sub <- dge_plot +
  labs(title = NULL)

library(patchwork)

combined_plot <-
  (pca_sub | dge_sub) +
  plot_layout(widths = c(2,1)) +
  plot_annotation(
  title = "Which Human Cell Model Best Reflects Infected COVID-19 Lung Tissue?",
  subtitle = "Comparison of SARS-CoV-2 infected cell lines and human lung tissue",
  caption = paste(
    "PCA was computed using the top 1000 genes with the highest variance across samples.",
    "",
    "Conclusion: The Calu-3 cell line infected with SARS-CoV-2 shows gene expression changes most similar to real COVID-19 human lung tissue.",
    "",
    "Dataset: GSE147507",
    "Analysis by: Ashley Batugo",
    sep = "\n"
  )
)

combined_plot
```

```{r}
ggsave(
  "lab7_results.png",
  combined_plot,
  width = 14,
  height = 7,
  dpi = 300
)
```


⸻

STEP 5 — Final interpretation

Summarize:
	•	Which model had highest correlation?
	•	Which had most shared DEGs?
	•	Which had strongest shared pathways?
	•	Which looked closest in PCA?

Your conclusion will likely be:

Calu-3 best matches COVID lungs → best model.

⸻

📌 Expected Outcome

You will find:
	•	Calu-3 ≈ most realistic
	•	NHBE ≈ good
	•	A549-ACE2 ≈ partially realistic
	•	A549 ≈ not realistic

This matches known biology and gives you a strong, clean narrative.